Open source landscape ecology tools

### 2020-12-12 paper/paper.Rmd
:pencil: Work on introduction and overview of R packages

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Author 3
  address: Adress author 3
  email: E-Mail author 3
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{pdflscape}
  \usepackage{multirow}
  \usepackage{placeins}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options: 
  chunk_output_type: console
---

```{r setup R, include = FALSE}
# load packages
library(dplyr)
library(forcats)
library(ggplot2)
library(googlesheets4)
library(patchwork)
library(stringr)
library(tibble)
library(tidyr)

@@

results <- readRDS("results.rds")

@@

@@

<!-- MH: Not super happy with the list i) - iv) -->

Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa) and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}. 
to this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}. 
Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}. 
Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze and visualize spatial data.

## Open-source software and R {#sec:open_source}

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}. 
Open-source software includes all software which is released under a license that allows to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the methodology \cite{Prlic2012,Powers2019}. 
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} and to customize existing software to their specific needs \cite{Steiniger2009}. 
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to many quite expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

One example of a successful open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}. 
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}. 
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task View: Analysis of Spatial Data* \cite{Bivand2019a} and *CRAN Task View: Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total. 
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even much higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on tools implemented in the *R* programming language.
The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
Because additionally to the handling of spatial data, also further task such as statistical modeling, the preparation of publication-ready figures and even the preparation of complete reports using *RMarkdown* \cite{Xie2018} can be done in the same working environment, the *R* programming language can be a very powerful tool.
Thus, in the first part we are give a comprehensive overview over existing *R* packages for landscape ecology. 
In the second part, we identify topics for which *R* packages are currently missing to the best of our knowledge. 
For more general overviews see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

@@

There are two major platforms to host *R* packages and make them easily accessible to potential users.
Firstly, hosting the package on *GitHub* allows users to install it with only one line of code. 
Additionally, hosting the package on *GitHub* provides many useful tools to communicate between developers and users, such as bug reports \cite{Wickham2015}.
Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides large visibility to the community, ease of installation and a certain quality standard \cite{Wickham2015}.
However, at the same time, releasing a *R* package on *CRAN* requires more initial work for the developers compared to *GitHub*.
Since most *R* packages are developed and maintained by the community, this is a good example of how open-source software development can facilitate innovation, reproducibility and reuse of code, especially since now many *R* packages depend on each other (i.e. package *x* requires and uses code from package *y*) \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
For more information about *R* package development, see \cite{Wickham2015}.

<!-- MH: Maybe we need to think about a table with the topics and the packages as a nice overview. Additionally, a figure where we show a typical workflow would be nice maybe -->
<!-- JN: can you think of one/a few typical workflows? -->

<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->
<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write something more about issues with archived packages in the second part (what's missing in R)... -->

@@

Most of the existing spatial data belong to one of two data models: spatial raster and spatial vectors.
In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The vector data model consists of two main elements, geometries, such as points, lines, polygons, and attributes.
Each geometry is connected to a row in an attribute table.
<!-- ... -->
A few hundred of different spatial data file formats were developed to store data in these formats.

While R has several build-in data structures, including data frames, it has not got any internal support for reading, processing, or visualizing spatial data.
However, there was a notable interest in spatial data analysis since the inception of this language, which is shown in a long R history of packages supporting spatial data (\cite{Lovelace2019}, page 10).
Comparing to dedicated GIS software (e.g., QGIS, ArcGIS), both raster and vector models have several implementations in the R language.
<!-- are pros/cons of that needed? -->
<!-- info about GDAL/PROJ? -->

The most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
Alternatively, *stars* package can be used to read and process raster data \cite{Pebesma2019}.
It is interconnected with the *sf* package, and beside raster support, it provides tools for spatial data cubes.

The *sp* package was a standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 R packages directly depends or imports *sp*.
However, it is now in a maintenance mode only with a recommended successor - the *sf* package.
<!-- terra new vector representations -->

Importantly, all of the above packages use the external libraries GDAL and PROJ, which allow for reading and writing of hundreds of spatial data formats and coordinates transformation.
They also have dedicated methods or functions to convert between representations, for example between *sp* and *sf* objects or *raster* and *stars* objects.
This can be useful when a given function expects different data representation than currently existing.

@@

Nowadays, spatial data for many spatial scales is available from abundance of sources.
Some of them require registration or having appropriate credentials and allows downloading only from the web interface.
However, many of the sources made their data publicly available either as a direct download or through an API connection.
This is used by some R packages allowing for downloading various environmental spatial dataset.

A plethora of land cover products on global, continental, and country scales has been developed, however, most of them cannot be directly downloaded from R. 
The exception is the National Land Cover Database (NLCD) data for the USA, which can be accessed through the *FedData* package \cite{R-FedData}.
The *elevatr* package provides a synthesis raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}.
It uses different elevation products for a given area and selected level of details (zoom/resolution).
<!-- how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->
The *rgbif* gives access to data from the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}.
The `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim version 1.4, while the new in-development package *geodata*<!-- https://github.com/rspatial/geodata/blob/master/R/worldclim.R --> provides an access to WorldClim version 2 data.
Additionally, the *rgee package* \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.
<!-- it is worth to search for more as there are a lot of changes going on -->

<!--JN: Max, do you want to add something to this list?-->

@@

As we mentioned in the previous sections, gathered or obtained data can be represented by different data types and stored in various file formats.
<!-- and read into R using one of the packages from Section \@ref(sec:spatrep). -->
However, several potential obstacles to applying this data for ecological analysis still exist.

The main issue for many landscape ecology studies is related to projections of the used data.
It is not only required to have all of the used datasets in the same projection, but also selection of a proper projection is crucial.
Coordinates in spatial data can represent one of many coordinate reference systems.
Two main groups of coordinate reference systems (CRS) exist, namely geographical and projected, each having many members.
In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
<!---->
However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
<!-- Projected CRS  -->
Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
<!-- Where to find advice on how to select projections-->
Gladly, all of the packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.

Second possible issue with spatial data is when the obtained data extends over a larger area than the study region.
Then, some spatial preprecessing, such as vector clipping or raster cropping, is needed.
Packages from Section \@ref(sec:spatrep) also make these operations possible.
Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications.

<!--MH: Short paragraph about scale; *grainchanger*; DOI: 10.1111/2041-210X.13177 -->
<!-- MH: landscapetools; DOI: 10.1111/2041-210X.13076 -->

Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass}<!--ADD R.LI CITATION-->, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R-->
Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.

<!--JN: should we mention ArcGIS here and some of its modules?-->
<!--JN: in the same vain as above: should we mention Google Earth Engine?-->

@@

Creating maps is essential when working with spatial data.
It has a role in checking the quality of data<!--(both coordinates and values)-->, data exploration, and finally in communicating the results.
<!-- role of making maps - from checking data, through the data exploration, to presenting the results  -->
<!-- maps in the computer age -->
Two main types of maps, static maps with full creator control and interactive maps where more control is given to the user, can be created in R.

All of the packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the `plot()` function.
However, their defaults are focused on quick visual inspection of the data rather than the creation of complete maps.
Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018}.
This package aims for creating publication-ready maps.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}.<!--this sentence can be easily removed-->
R's general-purpose plotting package *ggplot2* together with its extension *ggspatial* also offers some capabilities for plotting spatial data \cite{Wickham2016a,R-ggspatial}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}. 
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.

Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}. 
On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}. 
Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}). 
Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}. 
In graph theory  \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}.
The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are another tool to model functional connectivity. 
The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}. 
Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}. 
Such insights improve our understanding of metapopulation dynamics, speciation, species’ distributions and conservation \cite{Storfer2007}. 
By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}. 
As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.

<!-- MH: Work-in-Progress -->
<!-- MH: General population genetics -->
<!-- MH: Isolation by resistance  -->
<!-- MH: Bayesian Genetic Clustering -->
<!-- MH: Multivariate Ordination Techniques -->

@@

````{r packages, include =FALSE}

>>>>

---
title: Open source landscape ecology tools
# subtitle: Do you have a subtitle? If so, write it here
# titlerunning: Open source landscape ecology tools
# authorrunning: Short form of author list if too long for running head
thanks: | 
    Grants or other notes about the article that should go on the front 
    page should be placed here. General acknowledgments should be placed at the
    end of the article.
authors: 
- name: Maximillian H.K. Hesselbarth
  address: Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: ...
  address: ...
  email: ...
keywords:
- keyword 1
- keyword 2
- keyword 3
- keyword 4
- ...
abstract: |
  max. (200 words) Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract Abstract
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: | 
  \usepackage{caption}
  \usepackage{ltablex}
  \usepackage{pdflscape}
  \usepackage{multirow}
  \usepackage{placeins}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r setup R, include = FALSE}
# load packages
library(dplyr)
library(forcats)
library(ggplot2)
library(googlesheets4)
library(knitr)
library(patchwork)
library(stringr)
library(tibble)
library(tidyr)

@@

results <- readRDS("data/results.rds")

@@

<!-- MH: We really need to make sure, we introduce all technical terms well enough throughout the paper -->

@@

Landscape ecology focuses on how ecological processes are influenced and modified by the heterogeneous landscapes they occur in and simultaneously how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
In this context, landscape ecology considers, besides others, i) spatial and temporal dynamics of heterogeneous landscapes, ii) interactions, fluxes and exchange within these landscapes, iii) how the landscapes influence ecological processes (and vice versa), and lastly iv) how to manage these heterogeneous landscapes \cite{Risser1984,Turner1989}.

@@

Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995}, spatial context is important and ecological processes will vary spatially \cite{With2019}.
To this, the importance of scale was already raised decades ago \cite{Wiens1989,Levin1992,Jelinski1996} and is still of relevance until today \cite{Simova2012,Estes2018}.
Thus, in contrast to many other sub-disciplines of ecology, landscape ecology emphasizes especially spatial patterns to a high degree \cite{Risser1984}.
Consequently, the field of landscape ecology relies on software to preprocess, modify, model, analyze, and visualize spatial data.

## Open-source software and *R* {#sec:open_source}

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software which is released under licenses that allow to freely use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, and reliability as well as longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, the use of open-source software facilitates (computational) reproducibility and can allow a better understanding of the used methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Last but not least, even though not strictly necessary by definition \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost which allows to use the software without any costs in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.

One successful example of an open-source project is the *R* programming language and its extensions (also called packages) \cite{RCoreTeam2019}.
Firstly introduced in 1995 \cite{Smith2016}, today the programming language is among the most popular programming languages, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a growing body of packages designed to analyze spatial data subsequently emerged for the *R* programming language \cite{Bivand2006}.
This is documented by the growing *CRAN Task Views: Analysis of Spatial Data* \cite{Bivand2019a} and *Handling and Analyzing Spatio-Temporal Data* \cite{Pebesma2020} that currently list almost 300 packages in total.
Since the task views are maintained manually by just a few people, the actual amount of *R* packages related to spatial data is most likely even higher.
The growing popularity of the *R* programming language for spatial data analysis and landscape ecology can also be seen by the increasing number of related textbooks \cite{Fletcher2019,Lovelace2019,Pebesma2019a}.

The growing body of *R* packages related to spatial data processing and analyzing results in a high capability of the *R* programming language for landscape ecology.
Even though many other open-source tools for landscape ecology exist \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017}, in this review we are going to focus on software implemented in the *R* programming language.
For more general overviews of open-source software for landscape ecology see e.g. \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
Because additionally to handling spatial data, also further tasks such as statistical modeling, preparation of publication-ready figures and even preparation of complete reports can be done in the same working environment, the *R* programming language can be a very powerful tool.
Thus, in the first part we present an overview over existing *R* packages for landscape ecology.
In the second part, we present a survey in which we asked the community how they currently use the *R* programming language and to identify topics for which *R* packages are currently missing for landscape ecology.

@@

Since most *R* packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
There are two major online platforms to host *R* packages and make them accessible to potential users.
Firstly, *GitHub* allows to host source code under version control and users can install packages with one line using the *remotes* \cite{Hester2020} package.
Additionally, hosting the package on *GitHub* provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
Secondly, the *Comprehensive R Archive Network* (*CRAN*) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
\cite{Wickham2015}.
For developers, the guaranteed technical quality standard on *CRAN* requires more initial work for compared to *GitHub*, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard on *CRAN* also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of *R* packages, i.e. package *x* requires and uses code from package *y* \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
For more information about *R* package development, see \cite{Wickham2015}.

```{r fig-workflow, echo = FALSE, out.width = "100%", out.height = "65%", fig.align="center", fig.cap = "Exemplary worklow of spatial data analyses for landscape ecology using the R programming language. For all major tasks, a few example *R* packages are listed. Of course, many more packages for the specific tasks exist."}
include_graphics("data/Workflow.pdf")
```

<!-- MH: What about packages that are not on CRAN anymore, e.g. adehabitat? -->

<!-- JN: if you think this package (or similar) are still useful, maybe we can mention them somewhere is the first part of the paper with annotation about their state, and then also write something more about issues with archived packages in the second part (what's missing in R)... -->

<!-- MH: I would argue to only include packages that are actively maintained. It's already such a huge amount and I guess there is a reason why the packages were archived. Doesnt speak for their popularity -->

@@

<!-- MH: Make sure all libraries etc. have reference -->

While *R* has several build-in data structures, including vectors, matrices or lists, the base *R* programming language has no internal support for reading, processing, or visualizing spatial data.
However, because there is a notable interest in spatial data analysis, support for spatial data is provided by many *R* packages (\cite{Lovelace2019}, page 10).
Most spatial data belong to one of two contrasting data models, namely spatial raster and spatial vectors, and both data models have several implementations in the *R* language.
Importantly, basically all *R* packages for spatial data use the external *GDAL* and *PROJ* libraries, which allow for e.g. reading and writing of hundreds of spatial data formats, and coordinates transformation.
Additionally, *R*, *GDAL* and *PROJ* allow to convert between data models and specific implementations, which can be useful if given methods only exist for a specific data model or implementation.

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The probably most prominent package allowing for raster data representation is *raster* \cite{Hijmans2019}.
A *raster* successor aimed at the simpler interface and improved performance, *terra*, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
Alternatively, the *stars* package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.

The vector data model consists of two main elements i) geometries, such as points, lines, polygons, and ii) attributes and each geometry is connected to a value in an attribute table.
In many cases, this data model allows a more natural representation of landscape features, however, with the cost of higher computational demands <!--(add reference)-->.
The *sp* package was the standard for vector data representation for more than 10 years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 *R* packages directly depend or imports *sp*.
However, more recently *sp* will only be maintained (and not actively developed any more), and the recommended successor is the improved *sf* package \cite{Pebesma2018}.

<!-- JN: terra new vector representations -->

<!-- MH: I probably would not add terra here. As far as I see, the vector support in terra is basically just support to combine vector with raster data. The main focuse of the package is raster data though -->

@@

Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a *R* session.
Since public available data is becoming more and more prominent, so are *R* packages to accesses them.
Packages include the *FedData* package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the *elevatr* package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the *rgbif* to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, *rnaturalearth* to access the Natural Earth database \cite{South2017}, or *marmap* to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
Furthermore, the `get_data()` function from the *raster* package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package *geodata* <!--(add reference)--> provides an access to data from WorldClim v2.
Additionally, the *rgee* package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

<!-- JN: how to cite (or maybe we should just add a link to?) - https://github.com/tilezen/joerd/blob/master/docs/data-sources.md -->

<!-- MH: I dont really get it. For which package do we need this link? -->

@@

Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps <!-- (add reference) -->.
This is also referred to as the projection of the data and is often a first barrier in spatial data analysis.
It is not only required to have all of the used data in the same projection, but also selection of a proper projection is crucial.
This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion and different CRS are optimized for different regions of the world and scales <!-- (add reference) -->.
Coordinates in spatial data can represent one of many coordinate reference systems.
Two main groups of CRS exist, namely geographical and projected, each having many members.
In the first one, geographical, positions are specified by latitude and longitude coordinates in degrees.
However, in most landscape ecology studies, projected CRSs, which use some measurement units (e.g., meters), should be utilized.
Selected projected CRS should be based on projection that keeps important properties of spatial data intact (e.g., does not distort areas, shapes, distances, angles) and be appropriate for a given study area.
Tools to find an appropriate CRS for a certain region can be found at <https://spatialreference.org>, <https://epsg.org>, or <http://epsg.io>.
In used data have different CRS, all packages from Section \@ref(sec:spatrep) have interfaces for coordinates transformations.

A second common spatial data processing task is required when the available data extends over a larger area than the study region.
In this case, the pre-processing of spatial data can include vector clipping or raster cropping. 
Related to that, also masking of certain areas of the study region can be required by using e.g. certain spatial filters such as water bodies.
Packages from Section \@ref(sec:spatrep) also allow for these operations.
Additionally, they offer many other operations, such as merging or joining of spatial data, extraction of values from one dataset into another, raster resolution changes, or vector data simplifications. A comprehensive collection of methods to aggregate raster values to a coarser resolution can also be found in the *grainchanger* package \cite{Graham2019}. Furthermore, *landscapetools* is a collection of various utility functions for the raster data model.

Finally, there is a number of tools for landscape ecology implemented in GIS software, such as r.li or r.pi for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.<!--https://github.com/r-spatial/rgee/blob/examples/image/morphological_operations.R--> Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as *rgrass7* \cite{R-grass7} for GRASS GIS, *RSAGA* \cite{R-RSAGA} for SAGA GIS, and *rgee* \cite{R-rgee} for Google Earth Engine.

<!--JN: should we mention ArcGIS here and some of its modules?-->
<!-- MH: I don't think this is really important -->

<!--JN: in the same vain as above: should we mention Google Earth Engine?-->
<!-- MH: Again, I dont think thats really of relevance  -->

@@

Creating maps is essential when working with spatial data.
It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
*R* allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g. changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
Alternatively, the *mapview* package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
The *tmap* package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Both, *tmap* and *mapview* build upon the *leaflet* package and leaflet javascript library \cite{R-leaflet}. 
Also the popular plotting package *ggplot2* \cite{Wickham2016a} has an extension especially designed for plotting spatial data named *ggspatial* \cite{R-ggspatial}.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g. amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the *landscapemetrics* package (see \ref{sec:landscape_metrics}).
Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011} landscapes are described by nodes (i.e. habitat patches) connected by and functional connections called links \cite{Laita2011}.
The *grainscape* package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More general, the *igraph* packages \cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are another tool to model functional connectivity.
The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the *gdistance* package \cite{vanEtten2017}.
Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published *samc* package \cite{Marx2020}.

@@

Landscape genetics investigates how characteristics of landscapes interact with gene flow, genetic drift and selection \cite{Manel2003}.
Such insights improve our understanding of metapopulation dynamics, speciation, species' distributions and conservation \cite{Storfer2007}.
By explicitly including characteristics of landscape, landscape genetics is a way more realistic way of analysis than, e.g., metapopulation genetics \cite{Holderegger2006}.
As a result of its interdisciplinary, landscape genetics combines methods from different disciplines including landscape ecology, spatial statistics, geography and population genetics \cite{Storfer2007}.

<!-- MH: Work-in-Progress -->

<!-- MH: General population genetics -->

<!-- MH: Isolation by resistance  -->

<!-- MH: Bayesian Genetic Clustering -->

<!-- MH: Multivariate Ordination Techniques -->

@@

```{r packages, include =FALSE}
